Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-25 paper.Rmd
what

---
title: "Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
 - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. We are especially grateful to Paul Huth for his comments and guidance. All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (nd), who argue that their theory-based model can predict civil war onsets better than several atheoretical alternatives or a model with country-structural factors. We find that there are several important mistakes and that their key finding is entirely conditional on the use of parametrically smoothed ROC curves when calculating accuracy, rather than the standard empirical ROC curves that dominate the literature. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine. More importantly, we argue that rather than trying to contrast the roles of theory and \"atheoretical\" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant theories. 
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
header-includes:
  - \usepackage{floatrow}
  - \floatsetup[figure]{capposition=bottom}
  - \floatsetup[table]{capposition=top}
  - \usepackage[format=plain,labelfont=bf,up]{caption}
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

Although their analysis in the end does not support this point, we do not disagree that theory is important in predictive modeling. Of the various ways one could contrast ad hoc machine learning with "theoretical" modeling, B&S focus essentially on theory as a mechanism for variable selection. In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting. The problem is that the space of extant theories for a phenomenon like civil war is so broad and non-specific in terms of an exact specification of what variables and in which functional form should constitute a prediction model, that it's both not very clear what the "correct" model should be, nor easy to demonstrate that any particular "theoretical" model's ability to predict generalizes beyond the data that was used to create it in the first place. Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to "explain" a phenomenon to strengthen and whittle down the broad space of extant theories.

@@

The next three issues we encountered all concern information in B&S Table 2, which is the key table in which B&S report AUC-ROC values to demonstrate the "Escalation" models superiority.

@@

At this level of detail, the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical", with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

@@

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain. As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance. However, these comparisons must be made carefully to ensure those correct inferences are drawn. We continue to think that weak theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation. Social science has leaned on theory for (at least) seventy years, but it might be time to find useful ways to incorporate some of the recently developed methods in the arena of machine learning and modern statistics as well.

>>>>

---
title: "Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
 - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. We are especially grateful to Paul Huth for his comments and guidance. All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (nd), who argue that their theory-based model can predict civil war onsets better than several atheoretical alternatives or a model with country-structural factors. We find that there are several important mistakes and that their key finding is entirely conditional on the use of parametrically smoothed ROC curves when calculating accuracy, rather than the standard empirical ROC curves that dominate the literature. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine. More importantly, we argue that rather than trying to contrast the roles of theory and 
\"atheoretical\" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant predictive work.  Instead, we argue that predictive modeling and machine learning are effective tools for theory testing.

output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
header-includes:
  - \usepackage{floatrow}
  - \floatsetup[figure]{capposition=bottom}
  - \floatsetup[table]{capposition=top}
  - \usepackage[format=plain,labelfont=bf,up]{caption}
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

Although their analysis in the end does not support this point, we do not disagree that theory can important in predictive modeling. Of the various ways one could contrast ad hoc machine learning with "theoretical" modeling, B&S focus essentially on theory as a mechanism for variable selection. In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting. The problem is that the extent of extant theories for a phenomenon like civil war is so broad and non-specific in terms.  They do not provide much guidance on an exact specification of what variables and which functional form is required. As a result it is both not very clear what the "correct" model should be, nor easy to demonstrate that any particular "theoretical" model's ability to predict generalizes beyond the data that was used to create it in the first place. Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to "explain" a phenomenon to strengthen and whittle down the broad space of extant theories.

@@

The next three issues we encountered all concern information in B&S Table 2, which is the key table in which B&S report AUC-ROC values to demonstrate the "Escalation" model's superiority.

@@

Instantiating the theoretical model is more about choosing the correct variables. This is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical", with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

@@

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain. As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance. However, these comparisons must be made carefully to ensure those correct inferences are drawn. We continue to think that weak theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation. Social science has leaned on theory for (at least) seventy years, but it is time to find useful ways to incorporate recently developed methods in the arena of machine learning and modern statistics.
